---
title: "Group Document"
format: html
editor: visual
---

# Week 7

## s222803

New feature

```{r}
data <- 5
```

```{r}
data <- 2*2
data2 <- data * 2
```

## s223196

Fascinating technology

```{r}
1+1 == 2
```
### New feature
Here is a plot:
```{r}
print('No, there is not.')
```



## s223237

Hola

```{r}
library(tidyverse)
hola <- 9
```

## s222798

-   **T1:** Find the `GitHub` repository for the `ggplot2` R-package

    https://github.com/tidyverse/ggplot2

-   **Q1:** How many Contributors are there?

    295

-   **T1:** Find the `GitHub` repository for the `Linux kernel`

    https://github.com/torvalds/linux

-   **Q2:** How many Contributors are there?

    24643

-   **Q3:** Discuss in your group why having an organised approach to version control is central? And consider the simple contrast of the challenges you were facing when doing the course assignments.

<!-- -->

-   **Q4:** Again, find the `ggplot` GitHub site and see if you can find how many branches there are?

    71 branches

```{r}
1 + 1
```

### New feature

More things here..

```{r}
2*2
```
